Rethinking how students learn to design programs

### 2016-04-23 manuscript-body.tex
Add methods and context

>>>>

\section{Context and Methods}\label{context-and-methods}

\subsection{Context of the study}\label{context-of-the-study}

The data presented in this case are taken from an ongoing IRB-approved
study of undergraduate electrical engineering majors undertaken at
\emph{Flagship State}, a large, public research institution on the east
coast. For two semesters, I have followed a total of 10 students taking
``Intermediate Programming Concepts for Engineers.'' It is the second of
a required two-semester course sequence in programming.\footnote{Hereafter,
  I use ``Intermediate Programming'' to refer to the second course and
  ``Basic Programming'' to refer to the first.} Students can (and some
do) place out of Basic Programming via AP Computer Science credit, but
all Electrical and Computer Engineering students must take Intermediate
Programming.

Intermediate Programming has two 75-minute lectures per week and a
weekly discussion section led by an undergraduate Teaching Assistant
(TA). Typical enrollment is between 60 and 80 students per semester.
Like Basic Programming, Intermediate Programming is taught using the C
programming language, and it incorporates multi-week projects in C as
part of its assessment structure. Students must work individually on
four projects over the course of the semester, which together comprise
45\% of their final course grade.

Grading projects involves running students' compiled code against
automated tests that determine whether a student program's output
matches the instructor's canonical output. If a student's program
completely matches the canonical output, the student receives at least a
90\% grade on a project. The remaining 10\% are discretionarily
allocated ``style'' points, awarded for things like proper formatting,
code commenting, and functional decomposition (Field Notes).

This study centers on ``Flights Database,'' the second of four projects
assigned to Intermediate Programming students during the spring 2012
semester. Students were asked to build a text menu-based program that
would let users query information about airports and plan non-stop and
one-stop flights between airports. For this project, the instructor gave
students three separate text files as source material. The
\textbf{airports} file contained names of airports and their
three-letter abbreviation codes; the \textbf{routes} file contained
3-tuples of two airport codes and the route number of a flight flying
between them; the \textbf{flights} file contained a list of specific
flight information (including arrival and departure times) by route
number. Crucially, in order to be able to respond to user queries
students would need to build a program that could coordinate information
across all three files to return an answer.

My analysis details the work of Rebecca, a female first-year electrical
engineering major. I focus specifically on Rebecca's code for finding
``one-stop'' flights, which the instructor defined as ``all pairs of
ﬂights that route the user between the departure and arrival airports
with exactly 1 stop (i.e., a one-connection ﬂight)'' (Flights Database
handout, 2012). The one-stop problem is particularly challenging. To
solve it successfully students' code must accept a user's choice of
airports and day, then stitch together routes that involve two separate
flights in a way that passes stringent constraints for acceptable
layover times.

\subsection{Methods}\label{methods}

This study proceeds from an empirical challenge: how can we develop
accounts of students' programming activity that explain the form and
evolution of their code on a design project? The focused form of that
challenge for this study is ``how can we understand the unconventional
design choices embedded in Rebecca's one-stop flight code?'' To answer
that question, my study draws from three data streams: ethnographic
observation, clinical interviewing, and code snapshot analysis.

For two semesters, I ethnographically embedded myself in the same
instructor's section of Intermediate Programming. My aim throughout was
to see what students see in terms of course material, assignment
directives, and instruction. In fall 2011 I observed approximately 50\%
of the course lectures. I also independently completed all class
homeworks and three out of the four course projects to more fully
understand the course's assessments. In spring 2012 I continued
attending lectures, though less frequently, and began attending select
TA-led discussion sections. During both lectures and discussion sections
I took field notes while recording ambient audio using a LiveScribe
Pulse pen.

Rebecca was one of four students (three female, one male) willing and
able to participate in a series of 1-hour outside-of-class clinical
interviews during the spring 2012 semester.\footnote{In total, roughly
  30 students responded to my initial in-class solicitation to be
  contacted by email about my study. Of the students I emailed,
  approximately 8 students responded to my emails to schedule interview
  times. Of those 8, only four students were able to successfully find
  interview time slots that fit our respective schedules.} I interviewed
Rebecca five times, and in typical interviews I split time between
asking about her experiences in the course and giving her time in the
interview to work on her project code. During each interview, I
simultaneously used:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A Kodak Zi8 camera for video-recording our interactions
\item
  A LiveScribe Pulse pen to capture Rebecca's on-paper penstrokes
\item
  A MacBook Pro (early 2011) with screen-recording software to capture
  everything on-screen while Rebecca programmed
\end{enumerate}

The final component of data gathering is modeled after Jadud's (2006)
system for capturing students' code. My colleagues and I developed
software, built around the open-source version control system called
Git, that effectively creates an entire copy of a student's code---what
we call ``snapshots''---every time students invoke the compiler on their
code. Our software then sends those snapshots to a secure,
researcher-accessible server in real-time as they're created.
Consequently, I could plan each interview with Rebecca around
up-to-the-minute knowledge of her work---in some cases work she had
completed just hours before the interview---and tailor my interview
questions to emerging patterns in her code. In total, Rebecca's work
resulted in 958 compilation snapshots over the course of the semester.
